---
title: "RP tests"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
link-citations: TRUE
vignette: >
  %\VignetteIndexEntry{RP tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{RPtests}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Philosophy based on cross-validation

In chapter 7 of @ESL2009 the authors emphasize cross-validation estimates the expected prediction error (or generalization error) of a *method*. This is in contrast to estimating such errors for a *specific model*. There is a line of work that makes use of cross-validation and inherits its inferential focus on a method rather than a specific model chosen by that method [@wasserman2009; @meinshausen2009; @meinshausen2010; @shah2013] including the focus of this vignette which was proposed by @shah2018. This paper and its `R` implementation [-@rRPtests] roughly attempt to answer the question:

[**Does this method (typically) result in models with a good fit, or does it fail to capture signal that an alternate method could (typically) detect?**]{.smallcaps}

If the first method does fit well, then the residuals from that method should (typically) look like pure noise. Hence, if an alternate method can predict the residuals with good accuracy (better than noise) that would be evidence the first method is failing to capture some signal. Hence the name **RPtests**, based on residual prediction. The exact details depend on several (groups of) decisions:

1.  Probability modeling assumptions (foundational, and we will not question these here).$$
    \mathbf{y} = \mathbf{X} \beta + \sigma \epsilon
    $$

2.  The first method, or null method. We want to know if this method is good enough. *We will focus on using the lasso [-@tibshirani1996] to select variables* in a (potentially high-dimensional) linear model.

3.  One or more alternate methods that could be chosen depending on what we are concerned about, e.g. *whether we should include more predictor variables* or search for non-linear relationships.

4.  Analysis/pipeline choices including how to tune each method, estimating the variance of the noise, number of bootstrap samples, and so on (we will also not question these here).

We'll focus on one of the main cases investigated in that paper: testing a linear model against an alternative that includes additional predictors. To determine if the alternative predicts the residuals significantly better than noise the paper uses a parametric-residual bootstrap described below, but first we consider an important point about the dimension of the first (null) model.

## Low vs high-dimensions

In the low dimensional linear model case, the scaled residuals

$$
\hat{\mathbf{R}} = \frac{(\mathbf{I} - \mathbf{P})\mathbf{y}}{\|(\mathbf{I} - \mathbf{P})\mathbf{y}\|_2}
$$

are a pivotal quantity under the null hypothesis that the model fits (and $\mathbf P = \mathbf{X}(\mathbf{X}^T \mathbf{X})^{-1} \mathbf{X}^T$ is the projection or hat matrix). In this case the distribution of the residuals does not depend on any unknown parameters, so we can use that distribution for valid hypothesis tests or other inferences. This is the basis of the standard $F$-test for goodness-of-fit, the one reported at the end of a `summary(lm(y ~ x))`.

```{r }
set.seed(1)
x <- runif(50)
y <- x + rnorm(50)
summary(lm(y ~ x))
```

However, this does not work in the high-dimensional case, so @shah2018 prove some results using a combination of previous lasso theory and new technical results to conclude the residuals from lasso can be used in a similar way (with slight modifications) to construct tests. The theory relies on strong conditions sufficient for the lasso to estimate the support and signs of the true $\beta$ with high probability, but (as is typical in lasso theory literature) simulations show good empirical performance under more general settings where the strong sufficient conditions are relaxed.

We briefly note that @jankova2020 apply the same basic idea of predicting residuals to high-dimensional *generalized* linear models by using the Pearson residuals

$$
\hat R_i = \frac{Y_i - \mu(x_i^T \hat \beta)}{\sqrt{V(x_i^T \hat \beta)}}
$$

instead, where $\mu$ and $V$ are the conditional mean and variance functions, respectively.

### Lasso details

To calibrate the test, we need a distribution to compare the accuracy measure achieved by a null model. In the high-dimensional case the null model is assumed to be selected by lasso with cross-validation. The authors propose iterating the whole procedure multiple times to average over the randomness induced by cross-validation itself (note that makes more sense when our inferential focus is the *method* and not a *specific model* chosen by the method). Below is a restatement of Algorithm 1 in @shah2018, the parametric-residual bootstrap procedure for estimating a null distribution. For $b = 1, \ldots, B$:

1.  Replace $\mathbf{y}$ with $\mathbf{\hat y}$ from the original null model plus simulated residuals $\mathbf{\zeta}^{(b)} \sim N(0, \mathbf{I}_{n \times n})$. (Simulating a new response from the null model).\
2.  Fit sqrt-lasso [-@belloni2011] using the above simulated $\mathbf{y}^{(b)}$ and subtract new fit $\mathbf{\hat y}^{(b)}$. (Simulating residuals achieved by alternative method).\
3.  Normalize simulated residuals and compute some accuracy measure, such as RSS. (Obtain bootstrap distribution for testing).\

Note that the sqrt-lasso method can be replaced with other fitting procedures if we are interested in different alternatives, for example by using a tree based method if we are interested in predictive interactions.

Exercise: why not just predict normalized residuals $(\mathbf{y} - \mathbf{\hat y})/\hat \sigma$ and compare to accuracy in predicting $\mathbf{\zeta}^{(b)}$? 

## Simulations and examples

```{r setup}
library(RPtests)
library(hdi) # riboflavin dataset
```

We continue focusing on the high-dimensional linear model and as an alternative to a selected model we consider whether more variables should be included. @shah2018 do not consider selective inference, hence in that paper's simulations it seems there is no model selection involved in choosing the hypotheses. Type 1 error control is shown in simulations set up for the null to be true in all instances, rather than doing model selection and subsetting to instances where the null holds.

## References
